Visualizing Multi-way Contingency Tables

### 2005-11-18 trunk/vcd/paper/hsc.Rnw
modification to color chapter in HSC
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@500 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

The plots introduced in the previous section are basically composed of
tiles whose areas represent characteristics derived from the
contingency tables---observed and expected frequencies in the case of
mosaic and sieve plots, or residuals as visualized by association
plots. When using color for these tiles, it is imperative
to choose the right color palettes, derived from suitable color
spaces. Apart from esthetic considerations, wrongly chosen colors
might serously affect the analysis. For example,

@@

Now, software implementations are often based on the more convenient
Hue-Saturation-Value (HSV) colors (or the comparable Hue-Luminance-Saturation 
color scheme). Both spaces are rather similar transformations of the RGB space
\cite{vcd:Brewer:1999,vcd:Poynton:2000} and are very common implementations of
colors in many computer packages \cite{vcd:Moretti+Lyons:2002}. Each color in this space is
represented by three dimensions: the hue (dominant wavelength in the spectrum), 
the saturation (`colorfulness' or `pureness'), and 
the value (`brightness', amount of gray). These intuitive dimensions
make HSV colors easyier to specifiy than, e.g, RGB colors. 
But HSV colors have several disadvantages.
Most importantly, HSV colors are not perceptually
uniform because the three HSV dimensions map only poorly to the three
perceptual dimensions of the human visual system  \cite{vcd:Brewer:1999,vcd:Ihaka:2003}.
One important issue here is that HSV dimensions are confounded, e.g., saturation
is not uniform across different hues.
As an example, see Fig.~\ref{fig:colors} (left) 
showing a qualitative color palette in the HSV
space: although saturation and value are fixed, the fully saturated blue is
perceived much darker than the fully saturated red or green,
making it difficult to judge the size of shaded areas.
Furthermore, the flashy fully saturated HSV colors are hard 
to look at for a longer time. For similar reasons, it is equally difficult
to derive acceptable diverging palettes from the HSV space, i.e.,
bipolar scales containing colors ranging between two very distinct colors.
The upper part of Fig.~\ref{fig:diverging} shows a diverging
palette in the HSV space with colors ranging from a saturated red over a neutral white
to a saturated blue. Although the palette should be balanced with
respect to colorfulness and brightness, the red colors are perceived
to be more more intense and flashy than the corresponding blue
colors.

\setkeys{Gin}{width=\textwidth}
\begin{figure}
\begin{center}
<<colorfig,fig=TRUE,echo=FALSE,height=3,width=7,results=hide>>=
par(mfrow = c(1,2), mar = c(1,1,1,1), oma = c(0,0,0,0))
pie(rep(1,9), radius = 1, col = rainbow(9), labels = 360 * 0:8/9)
pie(rep(1,9), radius = 1, col = rainbow_hcl(9), labels = 360 * 0:8/9)
@ 
\caption{Qualitative color palette for the HSV (left) 
         and HCL (right) spaces.}
\label{fig:colors}
\end{center}
\end{figure}

@@

The use of colors that are more `in harmony' go back to Munsell 
\cite{vcd:Munsell:1905} %'
who introduced a notation for balanced colors. Based on those,
tools producing better palettes for specific tasks have been developed 
\cite{vcd:Harrower+Brewer:2003}. Other perceptually-based color
spaces, especially suited for computer displays, are the CIELAB and CIELUV
spaces \cite{vcd:CIE:2004} from which qualitative palettes for
statistical graphics have been derived \cite{vcd:Ihaka:2003}. A
transformation of the CIELUV space leads to the HCL (Hue-Chroma-Luminance) space.
Chroma loosely corresponds to colorfulness and luminance to
brightness, but in contrast to HSV colors, chroma is an
\emph{absolute} measure valid for all hues, and luminance can be
varied independently from the other two dimensions.
Qualitative color palettes can easily be obtained by holding constant
chroma and luminance, and dividing the hue range in equally-sized intervals. 
HCL colors with fixed chroma and luminance 
are always balanced towards the same grey 
and thus do not have the problem of varying saturations as the HSV
colors (see Fig.~\ref{fig:colors}, right). 
Similarly, diverging color palettes can be derived
\cite{vcd:Zeileis+Meyer+Hornik:2005} that are
suited, e.g., for visualizing residuals as we will see in the next
section. In principle, for each of the two bipolar hues of the
diverging palette, we could hold constant 
the luminance and only vary the chroma to get the different colors. 
However, it turns out that the resulting colors 
are too similar to be used for discriminating shaded
objects. Better contrasting can be achieved by \emph{simultaneously}
varying chroma and luminance, for example by choosing two points in
the chroma/luminance plane for a given hue and interpolating
linearly. For each half of the diverging palette, 
one of these points should be a ``neutral'' color,
and the other correspond to the ending part of the scale. 
The lower part of Fig.~\ref{fig:diverging} shows a diverging
palette using HCL colors. In contrast to the HSV counterpart, matching
colors (light red/blue and dark red/blue) 
are balanced to the same gray, and thus get the same perceptual ``weight''.

@@

All plots introduced in Sect.~\ref{sec:twoway} are composed of empty
tiles. It seems intuitive to used filled tiles to highlight information
of interest. Consider again the hospital example: in Fig.~\ref{fig:highlight},
we mark tiles for patients never or seldom visited to visualize their proportion in
the LOS strata. For optical clarity, we set the
spacing between the visit frequency tiles to 0. Clearly, the
proportion of these patients increases with LOS. In fact, such a
``stacked'' plot can also be interpreted as spine plot with highlighting
\cite{vcd:hummel:1996}, particularly useful in analyzing causal relationships in
categorical data with a binary dependent variable. More variations on
this theme, such as doubledecker plots, are treated in chapter XXX
(\textbf{REFERENCE TO OTHER CHAPTER(S)}).

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}
\begin{center}
<<highlight,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
mosaic(t(hospital), 
       mar = c(left = 3.5), 
       labeling_args = list(offset_labels = c(left = 0.5), 
         offset_varnames = c(left = 1, top = 0.5), set_labels = 
         list("Visit frequency" = c("Regular","Less than\nmonthly","Never"))),
       split = TRUE, spacing = spacing_highlighting, 
       gp = gpar(fill = c("gray","black","black")))
@ 
\caption{Spine plot with highlighting for the hospital data.}
\label{fig:highlight}
\end{center}
\end{figure}

>>>>

The plots introduced in the previous section are basically composed of
tiles whose areas represent characteristics derived from the
contingency tables---observed and expected frequencies in the case of
mosaic and sieve plots, or residuals as visualized by association
plots. When using color for these tiles, it is imperative
to choose the right color palettes, derived from suitable color
spaces. Apart from aesthetic considerations, wrongly chosen colors
might serously affect the analysis. For example,

@@

Now, software implementations are often based on the more convenient
Hue-Saturation-Value (HSV) colors (or the comparable Hue-Luminance-Saturation 
color scheme). Both spaces are rather similar transformations of the RGB space
\cite{vcd:Brewer:1999,vcd:Poynton:2000} and are very common implementations of
colors in many computer packages \cite{vcd:Moretti+Lyons:2002}. Each color in this space is
represented by three dimensions $(H, S, V)$: the hue $H$ (dominant wavelength in the spectrum, in $[0, 360]$), 
the saturation $S$ (`colorfulness' or `pureness', in $[0, 100]$), and 
the value $V$ (`brightness', amount of gray, in $[0, 100]$).\footnote{In many implementations,
all three dimensions are scaled to the unit interval instead of the coordinates used here.}
These intuitive dimensions
make HSV colors easyier to specifiy than, e.g, RGB colors. 
But HSV colors have several disadvantages.
Most importantly, HSV colors are not perceptually
uniform because the three HSV dimensions map only poorly to the three
perceptual dimensions of the human visual system  \cite{vcd:Brewer:1999,vcd:Ihaka:2003}.
One important issue here is that HSV dimensions are confounded, e.g., saturation
is not uniform across different hues.
As an example, see Fig.~\ref{fig:colors} (left) 
showing a qualitative color palette (colors $(H, 100, 100)$ for varying hues $H$) in the HSV
space: although saturation and value are fixed, the fully saturated blue is
perceived much darker than the fully saturated red or green,
making it difficult to judge the size of shaded areas.
Furthermore, the flashy fully saturated HSV colors are hard 
to look at for a longer time. For similar reasons, it is equally difficult
to derive acceptable diverging palettes from the HSV space, i.e.,
bipolar scales containing colors ranging between two very distinct colors.
The upper part of Fig.~\ref{fig:diverging} shows a diverging
palette in the HSV space with colors ranging from a saturated red $(0, 100, 100)$
over a neutral white $(H, 0, 100)$ to a saturated blue $(240, 100, 100)$.
Although the palette should be balanced with
respect to colorfulness and brightness, the red colors are perceived
to be more more intense and flashy than the corresponding blue
colors.

\setkeys{Gin}{width=\textwidth}
\begin{figure}
\begin{center}
<<colorfig,fig=TRUE,echo=FALSE,height=3,width=7,results=hide>>=
par(mfrow = c(1,2), mar = c(1,1,1,1), oma = c(0,0,0,0))
pie(rep(1,9), radius = 1, col = rainbow(9), labels = 360 * 0:8/9)
pie(rep(1,9), radius = 1, col = rainbow_hcl(9), labels = 360 * 0:8/9)
@ 
\caption{Qualitative color palette for the HSV (left) 
         and HCL (right) spaces. The HSV colors are $(H, 100, 100)$ and
	 the HCL colors $(H, 50, 70)$ for the same hues $H$.}
\label{fig:colors}
\end{center}
\end{figure}

@@

The use of colors that are more `in harmony' goes back to Munsell 
\cite{vcd:Munsell:1905} %'
who introduced a notation for balanced colors. Based on those,
tools producing better palettes for specific tasks have been developed 
\cite{vcd:Harrower+Brewer:2003}. Other perceptually-based color
spaces, especially suited for computer displays, are the CIELAB and CIELUV
spaces \cite{vcd:CIE:2004} from which qualitative palettes for
statistical graphics have been derived \cite{vcd:Ihaka:2003}. A
transformation of the CIELUV space leads to the HCL (Hue-Chroma-Luminance) space.
These colors are again specified by triplets $(H, C, L)$:
chroma $C$ loosely corresponds to colorfulness and luminance $L$ to
brightness, but in contrast to HSV colors, chroma is an
\emph{absolute} measure valid for all hues, and luminance can be
varied independently from the other two dimensions.
Qualitative color palettes can easily be obtained by holding constant
chroma and luminance, and using varying hues. HCL colors with fixed chroma and luminance 
are always balanced towards the same grey 
and thus do not have the problem of varying saturations as the HSV
colors (see Fig.~\ref{fig:colors}, right). 
Similarly, diverging HCL color palettes can be derived
\cite{vcd:Zeileis+Meyer+Hornik:2005} by interpolating again between a
neutral color such as $(H, 0, 90)$ and two colors with full chroma
such as blue $(260, 100, 50)$ and red $(0, 100, 50)$. The resulting
palette is shown in the lower part of Fig.~\ref{fig:diverging}. 
In contrast to the HSV counterpart, matching
colors (light red/blue and dark red/blue) 
are balanced to the same gray, and thus receive the same perceptual ``weight''.
Note, that this varies chroma and luminance are varied simultaneously, i.e.,
the full chroma colors are also darker than the neutral color. Of course,
it would also be possible to choose a darker neutral color (or lighter 
full chroma colors), however, by varying both chroma and luminace better 
contrasts can be achieved.

@@

All plots introduced in Sect.~\ref{sec:twoway} are composed of empty
tiles. It seems intuitive to use filled tiles to highlight information
of interest. Consider again the hospital example: in Fig.~\ref{fig:highlight},
we mark tiles for patients never or seldom visited to visualize their proportion in
the LOS strata. For optical clarity, we set the
spacing between the visit frequency tiles to 0. Clearly, the
proportion of these patients increases with LOS. In fact, such a
``stacked'' plot can also be interpreted as spine plot with highlighting
\cite{vcd:hummel:1996}, particularly useful in analyzing causal relationships in
categorical data with a binary dependent variable. More variations on
this theme, such as doubledecker plots, are treated in chapter XXX
(\textbf{REFERENCE TO OTHER CHAPTER(S)}).

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}
\begin{center}
<<highlight,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
mosaic(t(hospital), 
       mar = c(left = 3.5), 
       labeling_args = list(offset_labels = c(left = 0.5), 
         offset_varnames = c(left = 1, top = 0.5), set_labels = 
         list("Visit frequency" = c("Regular","Less than\nmonthly","Never"))),
       split = TRUE, spacing = spacing_highlighting, 
       gp = gpar(fill = rep(grey.colors(2)[2:1], 1:2)))
@ 
\caption{Spine plot with highlighting for the hospital data.}
\label{fig:highlight}
\end{center}
\end{figure}
